An Efficient Calculus for Generalised Parikh Images of Regular Languages

### 2022-07-05 main.tex
Intro

\usepackage{amsmath,empheq,fancybox}
\usepackage{paralist}
\usepackage{url}
\usepackage{color}
\usepackage{textcomp,listings}
\usepackage{array}
\usepackage{mymacros}
\usepackage{microtype}
\usepackage{listings}
\usepackage{csquotes}
\usepackage{proof}
\usepackage{cleveref}
\usepackage{algorithm2e}      
\usepackage{multirow}
\usepackage{mathpartir}
\usepackage{amsthm}
\usepackage{ebproof}
\usepackage{stackengine}

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  The image of the Parikh map is important in automata theory, offering a
  compact characterisation of an automaton. We contribute a novel understanding
  of how Parikh maps can be combined with arbitrary commutative monoid morphisms
  to efficiently represent real-world constraints found in String solvers.
  Furthermore, we show how this formulation can be efficiently implemented as a
  calculus that we call \Calculus{} in a theorem prover supporting Presburger
  logic. In particular, our calculus allows us to solve the common problem in
  string solvers of computing the Parikh image of products of arbitrarily many
  automata. We demonstrate this by implementing a tool called \Catra{}, which we
  evaluate on constraints produced by \Fudge{the PyEx benchmarks} when solved by
  the \Ostrich{} string constraint solver. We show that \Calculus{} strictly
  outperforms the standard eager approach described
  in~\citeauthor{generate-parikh-image} as well as the improved
  over-approximating method recently described
  by~\citeauthor{approximate-parikh}. \Calculus{} typically also outperforms
  the~\Nuxmv{} model checker, in particularly on unsatisfiable instances.
  Finally, \Fudge{\Calculus{} and therefore \Catra{} provides an efficient
  method for deriving the Presburger formula of a given Parikh image that
  outperforms any previously published method.}
\end{abstract}

@@

In summary, we contribute:
\begin{itemize}
\item The \Calculus{} calculus to efficiently compute (a homomorphism on) the
Parikh image of products of Parikh automata.
\item A specialised method for quantifier elimination that blends well with \Calculus{} to efficiently compute quantifier-free Presburger representations of a given Parikh image.
\item Experiments illustrating the performance of \Calculus{} on real-world examples from string solving, including \NrBenchmarks{} instances in a standardised format made available for future study.
\item The \Catra{} tool for solving such instances, containing an implementation of \Calculus{}, the over-approximation described in~\cite{approximate-parikh}, and an adapter for the~\Nuxmv{} model checker~\cite{nuxmv}.
\item Suggestions for how to efficiently implement \Calculus{} in a modern automated theorem prover, including strategies for case splitting, clause learning, and constraint propagation for connectedness.
\end{itemize}

\subsection{A demonstration of our approach}\label{sec:introduction:motivation}

To illustrate the key features of our approach, we start with two regular
languages, a commutative homomorphism $f$ on them that we care about, and a side
constraint for which we want to find values in $f$'s image over the intersection
of the two languages. We will use the same two regular languages, recognised by
two automata; $\SomethingCSomething{}$ (Figure~\ref{fig:something-c-something})
and $\AcaOrBc{}$ (Figure~\ref{fig:aca-or-bc}) for both examples but change the
constraint and homomorphism $f$.

\subsubsection{Counting string lengths}

@@

The na\"ive approach would be to first compute the product of the automata and
then compute the length of its product. However, we can take advantage of the
both constraints and definitions of $f$ to prune the automata before computing
the product and help reduce the blow-up. To give an intuition for the core idea
of \Calculus{}, we will annotate the transitions of $\SomethingCSomething$ and
$\AcaOrBc$ with variables illustrating how many times they are taken, and
perform informal reasoning on them corresponding to the reasoning performed by
our calculus.

@@

Note that we have already performed some reasoning in the starting figures -- we
use the same symbol for the number of times each of the upper transitions of
$\AcaOrBc$ are taken, since we know that they must be taken the same number of
times, and we introduced the constant $1$ for the middle transition in
$\SomethingCSomething$, since it must always be taken precisely once to reach
the accepting state. We use standard graph-theoretical flow reasoning to set the
initial flow into the starting state to 1 and use linear equations to enforce
balanced flow throughout the automaton to determine the relations between the
counting variables on each transition (e.g. $1 + {l_b}' = l_a + l_b$, or $l_a +
l_c' = 1$). Strictly speaking, this is where the constant number $1$ in
Figure~\ref{fig:something-c-something} comes: from $1 + r_\Sigma = x + r_\Sigma$
solved for $x$. Using analogous linear equation reasoning loses additional
variables and gives us the annotations in Figure~\ref{fig:example-simplify-1}.

@@

\subsubsection{Producing the Parikh image}\label{sec:introduction:parikh}

@@

Implementing such a calculus is straightforward in \Calculus, since the
definition of products was left out of the definition. All that would be needed
is an appropriate update of the definitions of products. Simiarly, \Catra was
written with modularity in mind, and it should be straightforward to extend both
the input grammar and automata implementations to accomodate multi-track
automata with arbitrary synchronisation.

\contents{\item I have no idea what to write here
\item multiple tracks}

>>>>

\usepackage{amsmath,empheq,fancybox}
\usepackage{paralist}
\usepackage{url}
\usepackage{color}
\usepackage{textcomp,listings}
\usepackage{array}
\usepackage{mymacros}
\usepackage{microtype}
\usepackage{listings}
\usepackage{csquotes}
\usepackage{proof}
\usepackage{cleveref}
\usepackage{algorithm2e}      
\usepackage{multirow}
\usepackage{mathpartir}
\usepackage{amsthm}
\usepackage{ebproof}

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  A common problem in string constraint solvers is computing the Parikh image, a
   set of linear equations that describe all possible combinations of character
   counts in strings of a given language. Automata-based string solvers
   frequently need to compute the Parikh image of large products (intersections)
   of automata, which in many operations is both prohibitively slow and
   memory-intensive. We contribute a novel understanding of how Parikh maps can
   be combined with arbitrary commutative monoid morphisms to represent
   real-world constraints on regular languages, most notably the length
   constraint. Furthermore, we show how this formulation can be efficiently
   implemented as a calculus, \Calculus{}, in an automated theorem prover
   supporting Presburger logic. We also provide an efficient method for deriving
   the Presburger formula of a given Parikh image, \Fudge{which is good for
   something} based on techniques for quantifier elimination. We implement
   \Calculus{} in a tool called \Catra{}, and evaluate it on constraints
   produced by the \Ostrich{} string constraint solver when solving
   \Fudge{standard string constraint benchmarks involving string lengths}. We
   show that our solution strictly outperforms the standard eager approach
   described in~\citeauthor{generate-parikh-image} as well as the
   over-approximating method recently described
   by~\citeauthor{approximate-parikh}, and in many cases also the~\Nuxmv{} model
   checker.
\end{abstract}

@@

In summary, we contribute:
\begin{itemize}
\item The \Calculus{} calculus to efficiently compute (a homomorphism on) the
Parikh image of products of Parikh automata.
\item A specialised method for quantifier elimination that blends well with \Calculus{} to efficiently compute quantifier-free Presburger representations of a given Parikh image. \Fudge{Why is this good?}
\item Experiments illustrating the performance of \Calculus{} on real-world examples from string solving, including \NrBenchmarks{} instances in a standardised format made available for future study.
\item The \Catra{} tool for solving such instances, containing an implementation of \Calculus{}, the over-approximation described in~\cite{approximate-parikh}, and an adapter for the~\Nuxmv{} model checker~\cite{nuxmv}.
\item Suggestions for how to efficiently implement \Calculus{} in a modern automated theorem prover, including strategies for case splitting, clause learning, and constraint propagation for connectedness.
\end{itemize}

\section{A demonstration of our approach}\label{sec:motivation}

To illustrate the key features of our approach, we start with two regular
languages, a commutative homomorphism $f$ on them that we care about, and a side
constraint for which we want to find values in $f$'s image over the intersection
of the two languages. We will use the same two regular languages,
$\mathtt{ac^*a|b(bb)^*c}$ and $\mathtt{.^*c.^*}$ for both examples but change
the constraint $\varphi$ and homomorphism $f$. In essence, we solve the
following set of constraints for all possible values of $s$ and
$\MonoidElement$:

\begin{equation*}
  \AndComp{}{\begin{aligned}
  & \mathtt{ac^*a|b(bb)^*c} \text{ accepts } s\\
  & \mathtt{.^*c.^*} \text{ accepts } s \\
  & \MonoidElement = f(s) \\
  & \varphi(\MonoidElement)
  \end{aligned}}
 \end{equation*}

We will represent the regular languages as the automata $\SomethingCSomething{}$
 (\cref{fig:something-c-something}) and $\AcaOrBc{}$ (\cref{fig:aca-or-bc})
 respectively.

\subsection{Counting string lengths}

@@

The na\"ive approach would be to first compute the product of the automata, then
use the standard approach to find its image under $f$, and then verify the
constraints. However, we can take advantage of the both constraints and
definitions of $f$ to prune the automata before computing the product and help
reduce the blow-up. To give an intuition for the core idea of \Calculus{}, we
will annotate the transitions of $\SomethingCSomething$ and $\AcaOrBc$ with
variables illustrating how many times they are taken, and perform informal
reasoning on them corresponding to the reasoning performed by our calculus.

@@

Note that we have already performed some reasoning in the starting figures -- we
use the same symbol for the number of times each of the upper transitions of
$\AcaOrBc$ are taken, since we know that they must be taken the same number of
times, and we introduced the constant $1$ for the middle transition in
$\SomethingCSomething$, since it must always be taken precisely once to reach
the accepting state. We use standard graph-theoretical flow reasoning to set the
initial flow into the starting state to 1 and use linear equations to enforce
balanced flow throughout the automaton to determine the relations between the
counting variables on each transition (e.g. $1 + {l_b}' = l_a + l_b$, or $l_a +
l_c' = 1$). Strictly speaking, this is where the constant number $1$ in
\cref{fig:something-c-something} comes: from $1 + r_\Sigma = x + r_\Sigma$
solved for $x$. Using analogous linear equation reasoning loses additional
variables and gives us the annotations in \cref{fig:example-simplify-1}.

@@

\subsection{Producing the Parikh image}\label{sec:introduction:parikh}

@@

Implementing such a calculus is straightforward in \Calculus, since the
definition of products was left out of the definition. All that would be needed
is an appropriate update of the definitions of products. Simiarly, \Catra was
written with modularity in mind, and it should be straightforward to extend both
the input grammar and automata implementations to accommodate multi-track
automata with arbitrary synchronisation.
